---
title: "Linear Algebra"
output: html_notebook
---

# Notebook for Linear Algebra course on Udacity

## Points and Vectors

* Points: represent location, written as (x,y,z, ...)
+ Cartesian coordinate: standard 2-dimensional 

* Vector: represent change in location, properties include 
magnitude and direction, written as [2,4]

# Vector Module

## Operating on Vectors

In Python, we can create vector objects to operate on 
vectors, however, R handles this much more naturally.

```{r}
my_vector <- c(1,2,3)
my_vector2 <- c(1,2,3)
my_vector3 <- c(-1,2,3)

print(my_vector == my_vector2)
print(my_vector == my_vector3)
```

Addition can be accomplished using + operator
```{r}
x <- c(1,2)
y <- c(4,3)

x + y
```

Subtraction is similar

```{r}
x <- c(1,3)
y <- c(5,1)
x - y
y - x
```

Scalar Multiplication scales the size of a vector. 

```{r}
2 * c(1,2)

-1/2 * c(3,2,1)
```

Scalar Quiz

```{r}
c(8.218, -9.341) + c(-1.129, 2.111) 
c(7.119, 8.215) - c(-8.223, 0.878)
7.41 * c(1.671, -1.012, -0.318)
```


## Magnitude and Direction

Magnitude is the square root of the sum of the square of a
vectors coordinates.

To define a direction of a vector, a vector of magnitude
1 is used (unit-vector). 

Normalization is the process of 
finding a unit vector in the same direction as a given vector.

2 step process: 

1. find its magnitude: $||\vec{v}|| = \sqrt{v^2_1 + v^2_2 + \dotsb + v^2_n}$
2. perform a scalar multiplication: $\vec{v} \times \frac{1}{||\vec{v}||}$ = unit vector in direction of v

Example: normalize $\vec{v} = [-1,1,1]$

Answer: 
$$ ||\vec{v}|| = \sqrt{(-1^2) + 1^2 + 1^2}= \sqrt3 $$
$$ [-1, 1, 1] \times \frac{1}{\sqrt{3}} = [\frac{-1}{\sqrt(3)}, \frac{1}{\sqrt(3)}, \frac{1}{\sqrt(3)}] $$
In R: 
```{r}
magnitude <- function(x) {
  sqrt(sum(x^2))
}
normalize <- function(x) {
  x / magnitude(x)
}
v <- c(-1, 1, 1)
normalize(v) 

magnitude(normalize(v))
```
## Inner Products (Dot Products)

Dot product of two vectors is the sum of the 
product of the corressponding components of both
vectors

$$ \vec{v} \cdot \vec{w} = v_1w_1 + v_2w_2 + \dotsb + v_nw_n$$
The angle between two vectors is found through the following method: 
$$ \theta = arccos\left(\frac {\vec{v} \cdot \vec{w}} {||v|| \cdot||w||}\right) $$
An interesting result of the above is that we can compare
the direction of vectors v and w such that if the dot product of v an w equal the product of their magnitudes, then v and w have the same direction. If they are equal ot the negative of the product of their magnitudes, then the angle between both vectors is 180 degrees (opposite direction). 

If the product is 0 and neither vw and w is 0, which means the two vectors are at a right-angle to each other. 

The magnitude of v is the square root of v times v. 

### Implementation 

```{r}
# Implement dot product & angle 
dot_product <- function(x,y,...) {
  sum(x * y)
  # Alternatively, x %*% y
}

theta <- function(x, y) {
  acos((dot_product(x,y)) / (magnitude(x) * magnitude(y)))
}

v <- c(1,5,7) 
w <- c(2,3,0)
dot_product(v, w)
theta(v, w)
```
### Quiz 

```{r}
dot_product(c(7.887, 4.138), c(-8.802, 6.776))
c(-5.955, -4.904, -1.874) %*% c(-4.496, -8.755, 7.103)

theta(c(3.183, -7.627), c(-2.668, 5.319)) 
theta(c(7.35, 0.221, 5.188), c(2.751, 8.259, 3.985)) * 180/pi
```
## Parallel and Orthogoanal Vectors

$\vec{v}$ and $\vec{w}$ are parallel if one is a scalar multiple of another.

$\vec{v}$ and $-2\vec{w}$ are thus parallel

$\vec{v}$ and $\vec{w}$ are orthogonal if their dot product is 0. 

$\vec{0}$ is parallel and orthogonal to all vectors, and  the only vector 
that is orthogonal to itself 

### Quiz

Check for parallelism and orthogonality

```{r}
is_parallel <- function(x, y) {
  # All equal alllows for small differences between numbers
  # Absolute is taken to allow for parallel but opposite direction
  all.equal(abs(normalize(x)), abs(normalize(y)))
}
is_parallel(c(-7.579, -7.88), c(22.737, 23.64))
is_parallel(c(-2.029, 9.97, 4.172), c(-9.231, -6.639, -7.245))
is_parallel(c(-2.328, -7.284, -1.214), c(-1.821, 1.072, -2.94))
is_parallel(c(2.118, 4.827), c(0,0))

all.equal(dot_product(c(-7.579, -7.88), c(22.737, 23.64)), 0)
all.equal(dot_product(c(-2.029, 9.97, 4.172), c(-9.231, -6.639, -7.245)), 0)
all.equal(dot_product(c(-2.328, -7.284, -1.214), c(-1.821, 1.072, -2.94)), 0)
all.equal(dot_product(c(2.118, 4.827), c(0,0)), 0)
```
## Projection 

v = v(||) + v(perp)

Computing projb(v): 
Assume theta < 90 degrees
Length of |v(parl)| = v dotporudct u
