---
title: "Linear Algebra"
output: html_notebook
---

# Notebook for Linear Algebra course on Udacity

## Points and Vectors

* Points: represent location, written as (x,y,z, ...)
+ Cartesian coordinate: standard 2-dimensional 

* Vector: represent change in location, properties include 
magnitude and direction, written as [2,4]

# Vector Module

## Operating on Vectors

In Python, we can create vector objects to operate on 
vectors, however, R handles this much more naturally.

```{r}
my_vector <- c(1,2,3)
my_vector2 <- c(1,2,3)
my_vector3 <- c(-1,2,3)

print(my_vector == my_vector2)
print(my_vector == my_vector3)
```

Addition can be accomplished using + operator
```{r}
x <- c(1,2)
y <- c(4,3)

x + y
```

Subtraction is similar

```{r}
x <- c(1,3)
y <- c(5,1)
x - y
y - x
```

Scalar Multiplication scales the size of a vector. 

```{r}
2 * c(1,2)

-1/2 * c(3,2,1)
```

Scalar Quiz

```{r}
c(8.218, -9.341) + c(-1.129, 2.111) 
c(7.119, 8.215) - c(-8.223, 0.878)
7.41 * c(1.671, -1.012, -0.318)
```


## Magnitude and Direction

Magnitude is the square root of the sum of the square of a
vectors coordinates.

To define a direction of a vector, a vector of magnitude
1 is used (unit-vector). 

Normalization is the process of 
finding a unit vector in the same direction as a given vector.

2 step process: 

1. find its magnitude: $||\vec{v}|| = \sqrt{v^2_1 + v^2_2 + ... + v^2_n}$
2. perform a scalar multiplication: $\vec{v} \times \frac{1}{||\vec{v}||}$ = unit vector in direction of v

Example: normalize $\vec{v} = [-1,1,1]$

Answer: 
$$ ||\vec{v}|| = \sqrt{(-1^2) + 1^2 + 1^2}= \sqrt3 $$
$$ [-1, 1, 1] \times \frac{1}{\sqrt{3}} = [\frac{-1}{\sqrt(3)}, \frac{1}{\sqrt(3)}, \frac{1}{\sqrt(3)}] $$
In R: 
```{r}
magnitude <- function(x) {
  sqrt(sum(x^2))
}
normalize <- function(x) {
  x / magnitude(x)
}
v <- c(-1, 1, 1)
normalize(v)

magnitude(normalize(v))
```

